---
title: "Stringy Index Simulation"
author: "Mukul Kumar"
output: html_document
---

# Introduction

The goal of this simulation is to explore the minimum and maximum values that might be observed for any 2D pattern using the stringy index available in the `tourr` package. The stringy index measures the linear structure of data by leveraging PCA eigenvalues, where higher values indicate stronger linearity.

This document implements a custom stringy index function and tests it against five distinct 2D patterns:
- **Uniform random points**
- **Clustered data**
- **Linear data with noise**
- **Perfectly linear data**
- **Circular data**

The simulation reports the generated data, the computed stringy index values, and the theoretical expectations for each pattern. The expected theoretical range for the stringy index is between **0.5** (equal variance in all directions) and **1.0** (all variance in one direction).

# Approach

To achieve the task, the following steps were implemented:
1. **Custom Stringy Index Function**: A function was created to compute the stringy index based on PCA eigenvalues. The index is calculated as the ratio of the first eigenvalue to the sum of all eigenvalues.
2. **Simulation of 2D Patterns**: Five distinct 2D patterns were generated to cover the full range of expected stringy index values.
3. **Validation and Reporting**: The simulation computes the stringy index for each pattern, reports the minimum and maximum observed values, and compares them with the theoretical expectations.

# Setup tourr and ggplot2

```{r setup, message=FALSE, warning=FALSE}
library(tourr)
library(ggplot2)
```

# Custom Stringy Index Function
The stringy index is computed using PCA eigenvalues. If the data has only one column, the index is set to 1 (all variance in one direction).
```{r stringy-index}
stringy_index <- function(data) {
  if(ncol(data) == 1) return(1)
  eig <- svd(cov(scale(data)))$d
  eig[1] / sum(eig)
}
```

# Simulating Stringy Index for Different 2D Patterns

The following function generates various 2D patterns and computes their stringy index. The patterns include:

->Uniform: Random points uniformly distributed in 2D space.
->Clustered: Two clusters of points with random noise.
->Linear: Points distributed along a line with added noise.
->Perfect Line: Points perfectly aligned along a straight line.
->Circular: Points arranged in a circular pattern.

```{r simulation}
simulate_stringy_range <- function(n = 100) {
  datasets <- list(
    uniform = matrix(runif(n * 2), ncol = 2),
    clustered = cbind(
      c(rnorm(n/2, mean = -3), rnorm(n/2, mean = 3)),
      rnorm(n)
    ),
    linear = cbind(1:n, 1:n + rnorm(n, sd = 2)),
    perfect = cbind(1:n, 1:n),
    circular = cbind(sin(seq(0, 2*pi, length.out = n)), 
                     cos(seq(0, 2*pi, length.out = n)))
  )
  
  results <- lapply(datasets, stringy_index)
  list(
    min = min(unlist(results)),
    max = max(unlist(results)),
    data = datasets,
    results = results
  )
}

set.seed(123)
sim_results <- simulate_stringy_range(n = 100)
```

# Results
The minimum and maximum stringy index values observed in the simulation are:

```{r results-table}
result_df <- data.frame(
  Pattern = names(sim_results$results),
  StringyIndex = unlist(sim_results$results)
)
result_df[order(-result_df$StringyIndex),]
```

# Theoretical Expectations

-The theoretical expectations for the stringy index values are as follows:

-Perfect line: 1.0 (all variance in one direction)
-Linear with noise: ~0.9 (most variance in one direction)
-Clustered data: ~0.6-0.7 (depends on the orientation of clusters)
-Circle/uniform noise: ~0.5 (equal variance in all directions)

The observed values from the simulation align well with these expectations, validating the implementation.

# Visualization of Patterns

The following scatter plots visualize the generated datasets along with their computed stringy index values:
```{r visualization}
par(mfrow=c(2,3), mar=c(4,4,2,1))
for (pattern in names(sim_results$data)) {
  data <- sim_results$data[[pattern]]
  index <- sim_results$results[[pattern]]
  plot(data, main=paste0(pattern, ": ", round(index, 3)), 
       xlab="X", ylab="Y", pch=19, cex=0.7)
}
par(mfrow=c(1,1))
```

# Conclusion
The simulation validates the stringy index by testing it on five distinct 2D patterns, with results aligning well with theoretical expectations. The index ranges from **0.5** (equal variance in all directions) to **1.0** (all variance in one direction), effectively capturing the linear structure of data.